---
  title: "Capstone Project Bloom Predictions"
author: "Amie Bui"
date: "2026-02-18"
output: pdf_document
---
  
  ```{r setup, include = FALSE, echo=FALSE}
# DO NOT ALTER THIS CHUNK
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  fig.width = 5,
  fig.asp = 0.618,
  dpi = 120,
  fig.align = "center",
  cache = FALSE
)
# Load required packages
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(nasapower))
```

# -----------------------United States: Import Data ----------------------------

```{r}
dc_weather <- read.csv("~/STAT490 Capstone/USW00013743.csv")
```

## Clean Data set

```{r}
dc_weather_clean <- dc_weather %>%
  mutate(
    Date = as.Date(Date),
    year = year(Date),
    TMEAN = (tmax + tmin) / 2
  ) %>%
  filter(!is.na(TMEAN))

head(dc_weather_clean)
```

## Bloom Dates

```{r}
bloom_url <- "https://raw.githubusercontent.com/GMU-CherryBlossomCompetition/peak-bloom-prediction/main/data/washingtondc.csv"
blooms <- read.csv(bloom_url)
```

## Calculate Chill Threshold

```{r}
chill_threshold <- 32


dc_weather_clean <- dc_weather_clean %>%
  mutate(
    Date = as.Date(Date),
    month_day = format(Date, "%m-%d"),
    #shift so chill corresponds with correct year
    bloom_year = year(Date) + if_else(month(Date) >= 11, 1L, 0L),
  )



chill_data <- dc_weather_clean %>%
  filter(month_day >= "11-01" | month_day <= "01-31") %>%
  group_by(bloom_year) %>%
  summarise(chill_days = sum(TMEAN < chill_threshold))

```

## Compute Growing Degree Days

i. If temp is greater than 45 degrees fahrenheit, it is counted as "warmth"

```{r}

gdd_data <- dc_weather_clean %>%
  filter(month_day >= "02-01" & month_day <= "04-30") %>%
  mutate(gdd = pmax(TMEAN - 45, 0)) %>%
  group_by(bloom_year) %>%
  summarise(total_heat = sum(gdd))

```

## Merge data

```{r}

merged_data <- chill_data %>%
  left_join(gdd_data, by = "bloom_year") %>%
  left_join(blooms, by = c("bloom_year" = "year"))%>% 
  select(-location, -lat, -long, - alt) %>%
  drop_na(bloom_doy, chill_days, total_heat)

```

## Split into test and training data

```{r}

merged_data <- merged_data %>% arrange(bloom_year)

n <- nrow(merged_data)
split_index <- floor(0.8 * n)

train_data <- merged_data[1:split_index, ]
test_data  <- merged_data[(split_index + 1):n, ]

```

## Fit Model

```{r}

model1 <- lm(bloom_doy ~ chill_days + total_heat + total_heat * chill_days,
             data = train_data)


```




## Testing on Held Out Test Set

```{r}

test_data$predicted_bloom <- predict(model1, newdata = test_data)
MAE <- mean(abs(test_data$bloom_doy - test_data$predicted_bloom))
MAE
```

# --------------------------Kyoto, Japan: Import Data Set-----------------------

## Get Weather Data for Kyoto Japan

```{r}

kyoto_weather <- get_power(
  community = "AG",
  temporal_api = "daily",
  lonlat = c(135.7681, 35.0116), #Location of Kyoto
  pars = c("T2M_MAX", "T2M_MIN"),
  dates = c("1981-01-01", "2023-12-31")
)

kyoto_weather <- kyoto_weather %>%
  mutate(TMEAN = (T2M_MAX + T2M_MIN) / 2)

```


## Bloom Dates Kyoto Japan

```{r}
kyoto <- read.csv("~/STAT490 Capstone/kyoto.csv")
```



## Calculate Chill Threshold

```{r}
chill_threshold <- 0


kyoto_weather <- kyoto_weather %>%
  mutate(
    #Shift chill over so it is associated with the correct bloom year
    bloom_year = if_else(MM >= 11, YEAR + 1L, YEAR)
  )


chill_data_japan <- kyoto_weather %>%
  filter(MM >= 11| MM <= 1) %>%
  group_by(bloom_year) %>%
  summarise(chill_days = sum(TMEAN < chill_threshold))

```

## Compute Growing Degree Days

i. If temp is greater than 45 degrees fahrenheit, it is counted as "warmth"

```{r}

gdd_data_japan <- kyoto_weather %>%
  filter(MM >= 2 & MM <=  4) %>%
  mutate(gdd = pmax(TMEAN - 7.222, 0)) %>%
  group_by(bloom_year) %>%
  summarise(total_heat = sum(gdd))

```

## Merge data

```{r}

merged_data_japan <- chill_data_japan %>%
  left_join(gdd_data_japan, by = "bloom_year") %>%
  left_join(kyoto, by = c("bloom_year" = "year"))%>% 
  select(-location, -lat, -long, - alt) %>%
  drop_na(bloom_doy, chill_days, total_heat)
```

## Split into test and training data

```{r}

merged_data_japan <- merged_data_japan %>% arrange(bloom_year)

n <- nrow(merged_data_japan)
split_index <- floor(0.8 * n)

train_data_japan <- merged_data_japan[1:split_index, ]
test_data_japan  <- merged_data_japan[(split_index + 1):n, ]

```

## Fit Model

```{r}

model2<- lm(bloom_doy ~ chill_days + total_heat + total_heat * chill_days,
            data = train_data_japan)

```


## Testing on Held Out Test Set

```{r}

test_data_japan$predicted_bloom <- predict(model2, newdata = test_data_japan)
MAE <- mean(abs(test_data_japan$bloom_doy - test_data_japan$predicted_bloom))
MAE
```
# ---------------------------Switzerland: Import Data Set-----------------------


## Get Weather Data for Liestal, Switzerland

```{r}

switzerland_weather <- get_power(
  community = "AG",
  temporal_api = "daily",
  lonlat = c(7.735, 47.484), #Location of Liestal
  pars = c("T2M_MAX", "T2M_MIN"),
  dates = c("1981-01-01", "2023-12-31")
)

switzerland_weather <- switzerland_weather %>%
  mutate(TMEAN = (T2M_MAX + T2M_MIN) / 2)

```


## Bloom Dates Liestal, Switzerland

```{r}
liestal <- read.csv("~/STAT490 Capstone/liestal.csv")
```



## Calculate Chill Threshold

```{r}
chill_threshold <- 0


switzerland_weather <- switzerland_weather %>%
  mutate(
    #Shift chill over so it is associated with the correct bloom year
    bloom_year = if_else(MM >= 11, YEAR + 1L, YEAR)
  )


chill_data_switzerland <- switzerland_weather %>%
  filter(MM >= 11| MM <= 1) %>%
  group_by(bloom_year) %>%
  summarise(chill_days = sum(TMEAN < chill_threshold))

```

## Compute Growing Degree Days

i. If temp is greater than 45 degrees fahrenheit, it is counted as "warmth"

```{r}

gdd_data_switzerland <- switzerland_weather %>%
  filter(MM >= 2 & MM <=  4) %>%
  mutate(gdd = pmax(TMEAN - 7.222, 0)) %>%
  group_by(bloom_year) %>%
  summarise(total_heat = sum(gdd))

```

## Merge data

```{r}

merged_data_switzerland <- chill_data_switzerland %>%
  left_join(gdd_data_switzerland, by = "bloom_year") %>%
  left_join(liestal, by = c("bloom_year" = "year"))%>% 
  select(-location, -lat, -long, - alt) %>%
  drop_na(bloom_doy, chill_days, total_heat)
```

## Split into test and training data

```{r}

merged_data_switzerland <- merged_data_switzerland %>% arrange(bloom_year)

n <- nrow(merged_data_japan)
split_index <- floor(0.8 * n)

train_data_switzerland <- merged_data_switzerland[1:split_index, ]
test_data_switzerland  <- merged_data_switzerland[(split_index + 1):n, ]

```

## Fit Model

```{r}

model3<- lm(bloom_doy ~ chill_days + total_heat + total_heat * chill_days,
            data = train_data_switzerland)

```


## Testing on Held Out Test Set

```{r}

test_data_switzerland$predicted_bloom <- predict(model3, newdata = test_data_japan)
MAE <- mean(abs(test_data_switzerland$bloom_doy - test_data_switzerland$predicted_bloom))
MAE
```

#--------------------United States Prediction 2026------------------------------

```{r}

accuweather_forecast_2026 <- read.csv("~/STAT490 Capstone/accuweather_forecast_2026.csv")

```

